---
title: "Assignment_5"
author: "Dan Rodrigues"
date: "26/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1 - Expectation and variance of a discrete random variable
Suppose that $α,β \in [0, 1]$ with $α+β ≤ 1$ and let $X$ be a discrete random variable with with distribution supported
on $\{0, 1, 5\}$. Suppose that $\mathbb{P} (X = 1) = α$ and $\mathbb{P}(X = 5) = β$ and $\mathbb{P} (X \not\in \{0, 1, 5\}) = 0$.

First, let's restate some important information:

- We have $\alpha, \beta \in [0,1]$, where $\alpha + \beta \leq 1$.
- We have $X$ being a discrete random variable, with possible outcomes being $\{0, 1, 5\}$.
- We also know that $\mathbb{P} (X = 1) = α$ and $\mathbb{P}(X = 5) = β$ and $\mathbb{P} (X \not\in \{0, 1, 5\}) = 0$.
- Hence, $\mathbb{P}(X=0) + \mathbb{P}(X=1) + \mathbb{P}(X=5) = 1$, and therefore $\mathbb{P}(X=5) = 1-(\alpha + \beta)$.

What is the probability mass function $p_X : \mathcal{S} → [0, 1]$ for $X$?
\[
p(x)=
  \begin{cases}
    1-\alpha-\beta  & \text{if $x$ = 0}\\
    \alpha          & \text{if $x$ = 1}\\
    \beta           & \text{if $x$ = 5}\\
    0               & \text{otherwise}\\
  \end{cases}
\]

What is the expectation of $X$?
\begin{aligned}
\mathbb{E}(X)&:=\Sigma_{x\in\mathbb{R}}x \cdot p_X(x) \\
  &= \alpha \cdot 1 + \beta \cdot 5 + (1-(\alpha + \beta)) \cdot 0 + \Sigma_{x\in\mathbb{R} \backslash \{0,1\}}x \cdot p_X(0) \\
  &= \alpha + 5 \beta
\end{aligned}

What is the variance of $X$?
\begin{aligned}
\text{Var}(X)&:=\mathbb{E}[{X-\mathbb{E}(X)}^2] \\
  &= \mathbb{E}(X^2) - \mathbb{E}(X)^2 \\
  &= (\alpha + 25\beta) - (\alpha + 5\beta)^2 \\
  &= \alpha + 25\beta - \alpha^2 -25\beta^2 - 10\alpha\beta
\end{aligned}

## 2 - Simulating data with the uniform distribution
```{r 2}
library(tidyverse)    # To load %>%, ggplot
set.seed(0)
n <- 1000
sample_X <- data.frame(U = runif(n)) %>%
  mutate(X = case_when(
    (0<=U) & (U<0.25) ~ 1,
    (0.25<=U) & (U<0.5) ~ 5,
    (0.5<=U) & (U<=1)~0)) %>%
  pull(X)

# We have P(0 ≤ U < 1/4) = 1/4, P(1/4 ≤ U < 1/2) = 1/4 and P(1/2 ≤ U < 1) = 1, so this code generates a random variable X with P (X = 1) = P (X = 5) = 1/4 and P (X = 0) = 1/2 as required.

sample_X_015 <- function(alpha, beta, n) {
  sample_X <- data.frame(U = runif(n)) %>%
    mutate(X = case_when(
      (0<=U) & (U<alpha) ~ 1,
      (alpha<=U) & (U<alpha+beta) ~ 5,
      (alpha+beta<=U) & (U<=1)~0)) %>%
    pull(X)
  return(sample_X)
}

alpha <- 1/2
beta <- 1/10
n <- 10000

# Simulated E(X)
mean(sample_X_015(alpha, beta, n))

# Calculated E(X)
alpha + 5 * beta

# The simulated and calculated expectation, E(X), are in agreement. This can be explained by the law of large numbers, which suggests that as the number of samples increases (tends to infinity) the sample average converges towards the expectation for sequences of independent and identically distributed random variables.

# Simulated Var(X)
sample_X <- sample_X_015(alpha, beta, n)
mean(sample_X)

# Calculated Var(X)
var(sample_X)

# Similar to the expectation, E(X), the sample variance is also close to the population variance, due to the large number of samples of i.i.d. random variables.

library(purrr)
set.seed(0)
n <- 100
alpha <- 1/10

simulation_by_beta <- data.frame(beta=seq(from=0, to=0.9, by=0.01)) %>%
  mutate(sample_X = map(.x=beta, ~sample_X_015(alpha, .x, n))) %>%
  mutate(sample_avg = map_dbl(.x=sample_X, .f=~mean(.x))) %>%
  select(-sample_X) %>%
  mutate(expectation=alpha+5*beta)

simulation_by_beta %>%
  head(5)

df_pivot <- simulation_by_beta %>%
  rename(Sample=sample_avg, Expectation=expectation) %>%
  pivot_longer(cols=!beta, names_to = "var", values_to = "val")

df_pivot %>%
  head(5)

df_pivot %>%
  ggplot(aes(x=beta, y=val, linetype=var)) +
    geom_line(data=df_pivot %>% filter(var=="Expectation")) +
    geom_point(data=df_pivot %>% filter(var=="Sample")) +
    labs(x="Beta", y="Mean", linetype="") +
    theme_bw()
```

## 3 - The Gaussian Distribution
```{r Gaussian-pdf}
x<-seq(-4,6,0.1)
normal_densities_by_x <- data.frame(x=x,density=dnorm(x,mean=1,sd=sqrt(1)), var=1)%>%
  rbind(data.frame(x=x ,density=dnorm(x, mean=1, sd=sqrt(2)), var=2))%>%
  rbind(data.frame(x=x, density=dnorm(x,mean=1, sd=sqrt(3)), var=3))

ggplot(normal_densities_by_x,aes(x, y=density, color=as.character(var), linetype=as.character(var))) +
  geom_line() +
  theme_bw() +
  labs(color="Variance",linetype="Variance",x="x",y="Density")
```

```{r Gaussian-cdf}
normal_cdf_by_x <- data.frame(x=x, cdf=pnorm(x, mean=1, sd=sqrt(1)), var=1) %>%
  rbind(data.frame(x=x, cdf=pnorm(x, mean=1, sd=sqrt(2)), var=2)) %>%
  rbind(data.frame(x=x, cdf=pnorm(x, mean=1, sd=sqrt(3)), var=3))

ggplot(normal_cdf_by_x, aes(x, y=cdf, color=as.character(var), linetype=as.character(var))) +
  geom_line() +
  theme_bw() +
  labs(color="Variance", linetype="Variance", x="x", y="Cumulative distribution function")
```

```{r Gaussian-Quantile}
probs=seq(0,1,0.01)

normal_cdf_by_x<-data.frame(p=probs,q=qnorm(probs,mean=1,sd=sqrt(1)),var=1)%>%
  rbind(data.frame(p=probs,q=qnorm(probs,mean=1,sd=sqrt(2)),var=2))%>%
  rbind(data.frame(p=probs,q=qnorm(probs,mean=1,sd=sqrt(3)),var=3))

ggplot(normal_cdf_by_x,aes(x=p,y=q,color=as.character(var),linetype=as.character(var)))+
  geom_line()+
  theme_bw()+
  labs(color="Variance",linetype="Variance",y="Quantile",x="Probability")
```

#### Note: remaining non-code questions will be omitted, and are included in the accompanying answer sheet

## 4 - The Binomial distribution and the Central Limit Theorem
```{r 4.1}
p<-0.7
n<-50

binom_df<-data.frame(x=seq(0,n)) %>%
  mutate(pmf=map_dbl(.x=x,~dbinom(x=.x,size=n,prob=p)))

binom_df%>%
  head(3)
```

```{r 4.2}
inc<-0.01
mu=n*p
sigma=sqrt(n*p*(1-p))

gaussian_df <- data.frame(x=seq(0,n,inc))%>%
  mutate(pdf=map_dbl(.x=x,~dnorm(x=.x,mean=mu,sd=sigma)))

gaussian_df%>%
    head(3)
```

```{r 4.3}
colors<-c("Gaussian pdf"="red", "Binomial pmf"="blue")
fill<-c("Gaussian pdf"="white", "Binomial pmf"="white")

ggplot() +
  labs(x="x",y="Probability") +
  theme_bw() +
  geom_line(data=gaussian_df, aes(x, y=pdf, color="Gaussian pdf"), size=2) +
  geom_col(data=binom_df, aes(x=x, y=pmf, color="Binomial pmf", fill="Binomial pmf")) +
  scale_color_manual(name = "", values=colors) +
  scale_fill_manual(name = "", values=fill) +
  xlim(c(20,50))
```

##5 - Exponential distribution
```{r 5.1}
my_cdf_exp<-function(x,lambda){
  if(x<0){
    return(0)
  }else{
    return(1-exp(-lambda*x))
  }
}

lambda <- 1/2
map_dbl(.x=seq(-1,4), .f=~my_cdf_exp(x=.x, lambda=lambda))
```

```{r 5.2}
test_inputs<-seq(-1,10,0.1)
my_cdf_output<-map_dbl(.x=test_inputs,.f=~my_cdf_exp(x=.x,lambda=lambda))
inbuilt_cdf_output<-map_dbl(.x=test_inputs,.f=~pexp(q=.x,rate=lambda))

all.equal(my_cdf_output,inbuilt_cdf_output)
```


```{r 5.3}
my_quantile_exp<-function(p,lambda){
  q <- -(1/lambda)*log(1-p)
  return(q)
}

inc<-0.01
test_inputs<-seq(inc,1-inc,inc)

my_quantile_output<-map_dbl(.x=test_inputs,.f=~my_quantile_exp(p=.x,lambda=lambda))
inbuilt_quantile_output<-map_dbl(.x=test_inputs,.f=~qexp(p=.x,rate=lambda))

all.equal(my_quantile_output,inbuilt_quantile_output)
```