---
title: "Assignment_1"
author: "Dan Rodrigues"
date: "28/09/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Task 1 - Create your first data frame
We create a data frame showing the combined information of animal names and their corresponding number of legs:
```{r Task_1}
animals <- c("Snake", "Ostrich", "Cat", "Spider") # Vector of animal names
num_legs <- c(0, 2, 4, 8) # Vector of number of legs
animals_df <- data.frame(animals, num_legs) # Generate data frame from two vectors of data
print.data.frame(animals_df) # Print out the created data frame
```

## Task 2 - Matrix Operations
We now perform some matrix operations. We start by generating three matrices from numeric sequences. We then use matrix operations to compute various outputs using these input matrices.
```{r Task_2}
# Create matrix X, which is 2x3 matrix of number descending from 12 to 2 in steps of -2
x_vect <- seq(12, 2, -2)
X <- matrix(data = x_vect, nrow = 2, ncol = 3)

# Create matrix Y, which is 2z2 matrix of numbers increasing from 1 to 4
Y <- matrix(data = seq(1, 4), nrow = 2, ncol = 2)

# Create matrix Z, which is 2x2 matrix o numbers increasing from 4 to 10 in steps of 2
Z <- matrix(data = seq(4, 10, 2), nrow = 2, ncol = 2)

# Compute matrix transposes of Y and Z
Y_t = t(Y)
Z_t = t(Z)

# Compute matrix sums. They should be the same (i.e. these operations are commutative)
sum_YZ <- Y + Z
sum_ZY <- Z + Y
check_for_equality_sum <- all((sum_YZ - sum_ZY) == 0)

# Compute matrix products. Note that these are NOT the same in both cases, as matrix multiplication is NON-COMMUTATIVE
prod_YZ <- Y %*% Z
prod_ZY <- Z %*% Y
check_for_equality_prod <- all((prod_YZ - prod_ZY) == 0)

# Compute element-wise (Hadarmard) product. This is commutative IF they have the same dimensions. Otherwise, not true, as can perform element-wise multiplication with all elements of the smaller matrix, but not so for the larger matrix
hprod_YZ <- Y * Z
hprod_ZY <- Z * Y
check_for_equality_hprod <- all((hprod_YZ - hprod_ZY) == 0)

hprod_YX <- Y %*% X
# hprod_XY <- X %*% Y
# Error in Y %*% X : non-conformable arrays
# This is due to incompatible matrix dimensions

# Find Y^-1:
Y_inv <- solve(Y)
Y_Y_inv <- Y %*% Y_inv
Y_inv_Y <- Y_inv %*% Y
# Note: although matrix multiplication is not generally commutative, Y_Y_inv and Y_inv_Y give equal results. This is because a matrix multiplied by its inverse will give the identity matrix, and vice versa, thereby resulting in the same output.

# Now compute Y_inv_X
Y_inv_X <- Y_inv %*% X

# Now we repeat this, but without pre-solving Y_inv
Y_inv_X_no_presolve <- solve(Y) %*% X
```

## Task 3 - Writing a Simple Function
We now produce a function to find the sum of all numbers divisible by 2 and / or 7 up to an input number n, and test it with two inputs, 14 and 1000.
```{r Task_3}
myFirstRFunc <- function (n) {
  # This function should take in a single numeric argument n, and output the sum of all those numbers strictly below n which are divisible by either 2, or 7, or both
  
  # Check / enforce that value is allowed... (error if negative number)
  sum_of_divisors <- 0
  for (i in 2:n-1) {
    # Check if divisible by 2 or 7
    if (i %% 2 == 0 | i %% 7 == 0) {
      sum_of_divisors <- sum_of_divisors + i
    }
  }
  return (sum_of_divisors)
}

print(myFirstRFunc(14)) # Should be 49
print(myFirstRFunc(1000)) # Should be 284787
```

## Task 4 - Making a wave plots
We now plot a wave (sine) function, given by the following:
$y=\sin(x)$

Alternatively, it can be represented using a Taylor function:
\[ \sin(x)=\sum_{n=1}^{\infty}(-1)^{n+1}\cdot \frac{c^{2n-1}}{(2n-1)!} \approx x-\frac{x^3}{3!}+\frac{x^5}{5!}-\frac{x^7}{7!}\ldots \]

```{r Wave Plots}
x <- seq(0, 20, 0.01)
y <- sin(x)
sin_df <- data.frame(x, y)
# print(head(sin_df, 3))
plot(x, y)
```